\bf Model Visualization Techniques for a Social Network Model

### 2018-08-23 paper-child.Rnw
re-write of section 5 complete.

The last principle of model visualization we use is exploring the process of fitting the model, instead of just focusing on the end result. This principle is perhaps the most important for SAOMs because the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user. Hiding the MCMC steps is practical and efficient if the goal is to fit a model to a set of longitudinal network data, obtain parameter estimates, and draw conclusions or make predictions. We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process instead of allowing them stay hidden.

One of the hidden steps in the SIENA method of moments algorithm is the ``microstep" process. A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{(m-1)}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$. One tie changes at a time until the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where $C$ is definined in Equation~\ref{eq:cdiff}.

This simulation process follows the steps of the continuous-time Markov chain. Each tie change in the CTMC is referred to as one ``microstep". At each microstep, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function. The options for change are removing a current tie, adding a new tie, or making no change at all. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of tie changes between two network observations. We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

@@

The first vizualization we present here is an animation of the simulated microsteps, when fitting model M1, that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz}. Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears. If there are no changes at a particular microstep, no changes are seen. Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}.

The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added. In both cases, the ego node is emphasized with color and size changes. When an edge is removed, the edge that currently exists is emphasized with the same color and size change as the node. In the animation, the edge shrinks and disappears as the ego node shrinks and changes color back to its original black. If an edge is being added, the red edge appears from nothing, grows large for emphasis, then changes color and size to match the rest of the edges.

In the network animation, we see the modeled steps of the unobserved CTMC process between two obsered networks. We see each part of the SAOM come into play: we first see the rate at which the nodes are selected to change, then we see the result of the actor maximizing its objective function by either deleting or adding a node. In addition, the layout of the nodes changes as edges are removed or added, which gives us a better sense of how the overall network structure changes with these individual tie changes.

We continue to use animation to view the changing structure of the network via adjacency matrix visualizations. Since there are 16 nodes in the data, numbered 1-16, we first use that order on the $x$ and $y$ axes for the matrix visualization. Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network, as shown in Figure~\ref{fig:adjmatviz}. This lack of perceived structure would contribute to further confusion as the network chages, so we adjust the ordering so that the viewer can better perceive the structure of the network. This process is known as matrix seriation \citep{seriation}.

@@

To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
\begin{equation}\label{eq:acum}
\mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). 
\end{equation}
We then performed a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation. In Figure~\ref{fig:adjmatorder}, we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$. Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.

We also use the PCA seriated layout to fix the layout in the animation of one of the microstep process simulations, some frames of which are shown in Figure~\ref{fig:adjmatani}. This animation is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process. Through this animation, which can be viewed at \url{https://vimeo.com/240092677}, we can see edges appearing, and then later on disappearing. These in-between steps are not shown when we look at the network at our discrete observation points, so by viewing this animation we can gain a better understanding of the underlying dynamics of this model.

<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps. (Ego and alter node labels are removed to declutter the figure.) Starting at frame 0, there are two clearly connected components: one in the bottom left corner, and one in the top right corner. By the end, the component in the bottom left has spread out and become sparsely connected, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
adjmatall <- matrix(0, nrow = 16, ncol = 16)
rownames(adjmatall) <- paste0("V", 1:16)
colnames(adjmatall) <- paste0("V", 1:16)
dat2 <- sfmsall %>% filter(rep == 1)
adjmatall2 <- NULL
for(i in 0:55){
  MS <- dat2[dat2$ms == i,c("from", "to")]
  myadjmat <- adjmatall
  for(j in 1:nrow(MS)){
    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
  }
  adjmatall2 <- rbind(adjmatall2, myadjmat)
}

@@

We also attempt to better understand the microstep process by visualizing the empirical transition probabilites for the first microstep in the process. These probabilities are defined in Equation~\ref{eq:pij}. We only do the first step of many because the \texttt{RSiena} transition probabilities for any two edges $i,j$ after the first step are only directly comparable for identical microsteps due to the conditioning on the current network state in the model. %By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$. Put more precisely, let the vector $X_i = (\{x_{ik}\}, \{x_{ki}\})_{k \neq i} \in \{0,1\}^{2(n-1)}$ represent the set of all tie variables involving node $i$ in the current network. For $p_{ij}$ to be comparable for any two network states $x$ and $x'$, the value of the vector $X_i$ must be the same in both states. It is possible to incorporate this information into our visualizations, but for now we look at only the first step, because we know without needing any complicated conditioning steps that the previous node state $X_i$ is identical in all cases. 
Thus we have 1,000 transition probabilities to examine: one transition probability for the first microstep taken for each of the simulations.
%Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
%In Figure~\ref{fig:histogramalters}, we present each ego node and the resulting probabilities of tie changes. The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made. The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change. In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected. There are, however, many steps that are never taken. Some ego nodes, like 1, 4, and 16, really explore the space of all possible outgoing ties. On the other hand, nodes like 13 and 15 explore very few possible outgoing ties.  It is unclear why this would be happening; it may have something to do with the ties in wave 1, where 13 and 15 have multiple connections, while 1 and 16 have none, but there are counterexamples of well connected nodes, such as 12, that explore the space more. In addition, we can see that the ties changed most often are removing ties, not adding ties. This matches the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.

<<heatmap, fig.align='center', fig.cap='An adjacency matrix visualization showing the empirical transition probabilities for the first microstep in 1,000 simulations. The ego node is on the y-axis, and the alter node is on the x-axis. There were many ties with empirical probability zero.', fig.height=2.5>>=

@@

# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
# 103, 3
ggplot(data = probsDat,
       aes(y = factor(ego, levels = rev(order1)),
             x = factor(alter, levels = order1),
             fill = prob)) +
  geom_tile(color = 'black') +
  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1)) +
  labs(y = "ego", x = "alter")+
  scale_x_discrete(position = 'top') +
  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
  ThemeNoNet + 
  theme(panel.grid = element_blank()) +
  coord_fixed()
@

%Another way to view these transition probabilities is through the adjacency matrix visualization. 
In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}. This adjacency matrix visualization shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations. The figure is noticeably sparse: %showing the lack of coverage in the model.
of the $16\cdot 15 = 240$ possible changes for the CTMC to make, only 103, or about 43\%, are taken in the 1,000 simulated chains. This leaves %reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken. This effect could only be exacerbated as more steps are taken in the CTMC, leading to 
a very large area of our possible network space unexplored by the first steps in the SAOM model fitting process. We present additional evidence for lack of coverage in Figure~\ref{fig:alledgests} the Appendix. The Figures~\ref{fig:heatmap} and \ref{fig:alledgests} show that the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism, a finding we were only able to discern thanks to model visualization techniques.

>>>>

When fitting a SAOM in \texttt{RSiena}, several simulation steps are hidden from the user. One of the hidden steps in the SIENA method of moments algorithm is the \textit{microstep} process. At each microstep, an \textit{ego} node is selected to make a change, and the chosen ego node randomly makes one change in its ties to maximize its objective function. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of differences between two network observations. We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain. Hiding the microsteps is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions. But in order to understand \textit{how} the models are fit, we need to extract and explore the different steps of the process. We use these steps to visually explore the algorithms, so that we are visualizing the process, and not just the main result.

%A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{(m-1)}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$. One tie changes at a time until the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where $C$ is definined in Equation~\ref{eq:cdiff}.

%This simulation process follows the steps of the continuous-time Markov chain. Each tie change in the CTMC is referred to as one ``microstep".  The options for change are removing a current tie, adding a new tie, or making no change at all.

@@

To visualize the underlying algorithm, we animate the microsteps that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz}. The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represent one possible path from wave 1 to wave 2 according to the model M1. We show each change in the network as follows: we first increase the size and change the color of the ego node to focus attention to its ties. Then, the changing tie either fades in or fades out. If there are no changes at a particular microstep, the microstep does not appear. Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}. Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. Frames 15-28 of Figure~\ref{fig:makinggif} show a deleted tie, while frames 71-84 show an added tie. For both changes, the ego node is emphasized with color and size. When a tie is removed, it is emphasized with the same color and size as the node. As the animation proceeds, the tie shrinks and disappears as the ego node shrinks and changes color back to black. If a tie is being added, the it appears red from nothing, grows large for emphasis, then changes color and size to match the rest of the edges. In this network animation, we see the modeled steps of the \textit{unobserved} CTMC process between two observed networks. We see each part of the SAOM come into play: we first see the selection of nodes to change via the rate function, then we see the result of the actor maximizing its objective function by either deleting or adding a node. Finally, the layout of the nodes changes as edges are removed or added, which demonstrates how the overall network structure changes with each of the individual tie changes. Combining the tie changes in this animation makes the CTMC process the focus and demonstrates the underlying change mechanism of the dynamic network according to the chosen model.

@@

We continue to use animation to view the changing structure of the network via adjacency matrix visualizations. In the first adjacency matrix visualization in Figure~\ref{fig:adjmatorder}, we use the node ID to order the rows and columns. This arbitrary ordering does not easily convey the underlying structure of the network. We use \textit{matrix seriation} to reorder the rows and columns of the adjacency matrix visualization so that the structure of the network is apparent \citep{seriation}. %This lack of perceived structure would contribute to further confusion as the network chages, so we adjust the ordering so that the viewer can better perceive the structure of the network. This process is known as matrix seriation 
To seriate the matrix, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
\begin{equation}\label{eq:acum}
\mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). 
\end{equation}
We then performed a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation. The resulting visualization for the first wave of data is shown in Figure~\ref{fig:adjmatorder}. %we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$. 
Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the visualization ordered by node ID. We continue to use the seriated matrix in another animation of the microstep process, which is avaiable at \url{https://vimeo.com/240092677}. A few frames of the video are shown in Figure~\ref{fig:adjmatani}. In the animation, a square appears or disappears as that edge appears or disappears in the microstep process. Through this animation, we can see edges appearing, and then later on disappearing. This visualization also shows the microsteps where the nodes do not make a change. From start to finish, we see network evolve according to the model: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row. The adjacency matrix animation shows the many ways the algorithm explores the data space. By viewing this animation, we can gain a better understanding of the dynamics of this model.

<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps. The node lables are removed to declutter the figure. Starting at frame 0, there are two clearly connected components: one in the bottom left corner, and one in the top right corner. By the end, the component in the bottom left has spread out and become sparsely connected, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
adjmatall <- matrix(0, nrow = 16, ncol = 16)
rownames(adjmatall) <- paste0("V", 1:16)
colnames(adjmatall) <- paste0("V", 1:16)
dat2 <- sfmsall %>% filter(rep == 1)
adjmatall2 <- NULL
for(i in 0:55){
  MS <- dat2[dat2$ms == i,c("from", "to")]
  myadjmat <- adjmatall
  for(j in 1:nrow(MS)){
    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
  }
  adjmatall2 <- rbind(adjmatall2, myadjmat)
}

@@

<<heatmap, fig.align='center', fig.cap='Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep observed in 1,000 simulations. The ego node is on the y-axis, and the alter node is on the x-axis. There were many ties with empirical probability zero.', fig.height=2.5>>=

@@

ansnullchains %>%
  filter(period == 1) %>%
  group_by(rep) %>%
  mutate(msno = row_number()) %>%
  filter(msno == 1) %>%
  ungroup() %>% group_by(ego, alter) %>% 
  tally() %>% ungroup() %>% mutate(empprob = n/1000) %>% 
  mutate(ego = as.numeric(as.character(ego))+1,
         alter = as.numeric(as.character(alter))+1) ->empprobsDat

# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
# 103, 3
p1 <- ggplot(data = probsDat,
       aes(y = factor(ego, levels = rev(order1)),
             x = factor(alter, levels = order1),
             fill = prob)) +
  geom_tile(color = 'black') +
  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1)) +
  labs(y = "ego", x = "alter")+
  scale_x_discrete(position = 'top') +
  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
  ThemeNoNet + 
  theme(panel.grid = element_blank()) +
  coord_fixed() + 
  ggtitle("Theoretical Probabilities")
p2 <- ggplot(data = empprobsDat,
       aes(y = factor(ego, levels = rev(order1)),
             x = factor(alter, levels = order1),
             fill = empprob)) +
  geom_tile(color = 'black') +
  scale_fill_gradient(low = 'white', high = 'black', name = "prob") +
  labs(y = "ego", x = "alter")+
  scale_x_discrete(position = 'top') +
  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
  ThemeNoNet + 
  theme(panel.grid = element_blank()) +
  coord_fixed() + 
  ggtitle("Empirical Probabilities")
gridExtra::grid.arrange(p1, p2, nrow = 1)
@

We also attempt to better understand the microstep process by visualizing the theoretical and observed transition probabilites for the first microstep in the process. These probabilities are defined in Equation~\ref{eq:pij}. We only do the first step of many because each microstep is dependent on the previous network state. %By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$. Put more precisely, let the vector $X_i = (\{x_{ik}\}, \{x_{ki}\})_{k \neq i} \in \{0,1\}^{2(n-1)}$ represent the set of all tie variables involving node $i$ in the current network. For $p_{ij}$ to be comparable for any two network states $x$ and $x'$, the value of the vector $X_i$ must be the same in both states. It is possible to incorporate this information into our visualizations, but for now we look at only the first step, because we know without needing any complicated conditioning steps that the previous node state $X_i$ is identical in all cases. 
There are 1,000 transitions to examine: the first microstep taken in each one of the simulations. %Another way to view these transition probabilities is through the adjacency matrix visualization. 
We use another seriated adjacency matrix visualization to show theoretical and empirical transition probabilities of the first microstep in Figure~\ref{fig:heatmap} for each tie that was the first change in one of the simulations. The figure is noticeably sparse: %showing the lack of coverage in the model.
of the $16\cdot 16 = 256$ possible choices for the CTMC to make, only 103, or about 43\%, are made in the 1,000 simulated chains. This leaves %reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken. This effect could only be exacerbated as more steps are taken in the CTMC, leading to 
a large area of data space unexplored by the first step in the SAOM model fitting process. We present additional evidence for lack of coverage in Figure~\ref{fig:alledgests} the Appendix. The Figures~\ref{fig:heatmap} and \ref{fig:alledgests} show that the model M1 and the the corresponding fitting process do not explore the data space enough to adequately capture the network change mechanism, a finding we were only able to discern thanks to model visualization techniques. 
%Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
%In Figure~\ref{fig:histogramalters}, we present each ego node and the resulting probabilities of tie changes. The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made. The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change. In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected. There are, however, many steps that are never taken. Some ego nodes, like 1, 4, and 16, really explore the space of all possible outgoing ties. On the other hand, nodes like 13 and 15 explore very few possible outgoing ties.  It is unclear why this would be happening; it may have something to do with the ties in wave 1, where 13 and 15 have multiple connections, while 1 and 16 have none, but there are counterexamples of well connected nodes, such as 12, that explore the space more. In addition, we can see that the ties changed most often are removing ties, not adding ties. This matches the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.
